---
title: "Mobility & migration in LATAM "
output: html_document
date: "2025-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300
)
```

```{r echo=FALSE, warning=FALSE}

#install.packages("factoextra")


# Cargar librerías
library(tibble)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(cluster)
library(factoextra)
library(tidyr)
library(kableExtra)
library(migest)
library(scales)

rm(list = ls())

##

```





##Indicadores globales

## ARG y CHL comparación



### GNI per Capita, PPP (constant international USD)



GNI per capita is the Gross National Income (total income earned by residents) divided by the total population, providing a measure of a country's average income and living standards. 


```{r echo=FALSE, warning=FALSE,fig.width=8, fig.height=6}

# Install and load required packages
if (!require(WDI)) install.packages("WDI")

library(WDI)


# Define Latin American countries for regional average
latin_america_codes <- c(
  "AR", "BO", "BR", "CL", "CO", "CR", "CU", "DO", "EC", "SV", 
  "GT", "HN", "MX", "NI", "PA", "PY", "PE", "UY", "VE"
)

# Paleta homogénea (del análisis ISDM)
colores_pais <- c(
  "Chile" = "#1BA3A3",       
  "Argentina" = "#D73027"    
)

colores_completos <- c(
  "Argentina" = "#D73027",
  "Chile" = "#1BA3A3",
  "Media Regional (América Latina)" = "#FFA726"
)

```

## GNI per Capita (PPP, Constant International USD)

```{r cho=FALSE,}


# Get GNI data for all Latin American countries
gni_regional_data <- WDI(
  country = latin_america_codes,
  indicator = "NY.GNP.PCAP.PP.CD",
  start = 1995,
  end = 2022
)

# Clean and prepare regional data
gni_regional_data <- gni_regional_data %>%
  rename(
    year = year,
    country = country,
    gni_ppp = NY.GNP.PCAP.PP.CD
  ) %>%
  filter(!is.na(gni_ppp))

# Calculate regional average by year
regional_average <- gni_regional_data %>%
  group_by(year) %>%
  summarise(
    gni_ppp = mean(gni_ppp, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA"  # Adding iso2c for consistency
  )

# Get data specifically for Argentina and Chile
gni_focus_countries <- gni_regional_data %>%
  filter(iso2c %in% c("AR", "CL"))

# Combine focus countries with regional average
gni_combined <- bind_rows(gni_focus_countries, regional_average)

# Create country labels for better display
gni_combined <- gni_combined %>%
  mutate(
    country_label = case_when(
      iso2c == "AR" ~ "Argentina",
      iso2c == "CL" ~ "Chile",
      country == "Regional Average" ~ "Media Regional (América Latina)",
      TRUE ~ country
    )
  )


# Plot the time series
ggplot(gni_combined, aes(x = year, y = gni_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
    title = "GNI per Capita (PPP, Constant International USD)",
    subtitle = "Argentina & Chile (1995–2022)",,
    x = "Year",
    y = "GNI per Capita (USD, PPP)",
    color = "Country",
    caption = "Source: World Bank (WDI), indicator NY.GNP.PCAP.PP.CD"
  ) +
  scale_color_manual(values = colores_completos)+
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))


# Plot the time series
pp<-ggplot(gni_combined, aes(x = year, y = gni_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
  #  title = "GNI per Capita (PPP, Constant International USD)",
  #  subtitle = "Argentina & Chile (1995–2022)",,
    x = "Year",
    y = "GNI per Capita (USD, PPP)",
    color = "Country"
  #  caption = "Source: World Bank (WDI), indicator NY.GNP.PCAP.PP.CD"
  ) +
   scale_color_manual(values = colores_completos)+
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))


#save plot
ggsave("output_context/gni_per_capita_arg_chl_media.png", plot = pp, width = 8, height = 6)

```


GNI per Capita (OTROS PAISES)



```{r cho=FALSE,}


# Get GNI data for all Latin American countries
gni_regional_data <- WDI(
  country = latin_america_codes,
  indicator = "NY.GNP.PCAP.PP.CD",
  start = 1995,
  end = 2022
)

# Clean and prepare regional data
gni_regional_data <- gni_regional_data %>%
  rename(
    year = year,
    country = country,
    gni_ppp = NY.GNP.PCAP.PP.CD
  ) %>%
  filter(!is.na(gni_ppp))

# Calculate regional average by year
regional_average <- gni_regional_data %>%
  group_by(year) %>%
  summarise(
    gni_ppp = mean(gni_ppp, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA"  # Adding iso2c for consistency
  )

# Get data specifically for Argentina and Chile
gni_focus_countries <- gni_regional_data %>%
  filter(iso2c %in% c("AR", "CL", "BR","UY"))

# Combine focus countries with regional average
gni_combined <- bind_rows(gni_focus_countries, regional_average)

# Create country labels for better display
gni_combined <- gni_combined %>%
  mutate(
    country_label = case_when(
      iso2c == "AR" ~ "Argentina",
      iso2c == "CL" ~ "Chile",
      iso2c == "BR" ~ "Brasil",
      iso2c == "UY" ~ "Uruguay",
      country == "Regional Average" ~ "Media Regional (América Latina)",
      TRUE ~ country
    )
  )


# Plot the time series
ggplot(gni_combined, aes(x = year, y = gni_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
    title = "GNI per Capita (PPP, Constant International USD)",
    x = "Year",
    y = "GNI per Capita (USD, PPP)",
    color = "Country",
    caption = "Source: World Bank (WDI), indicator NY.GNP.PCAP.PP.CD"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))




```


### GDP per Capita

GDP per capita is calculated as the value of all final goods and services produced each year in a country (the gross domestic product), divided by the population. It represents the average economic output per person



**GDP con media regional**


```{r echo=FALSE, warning=FALSE}


# Get GDP data for all Latin American countries
gdp_regional_data <- WDI(
  country = latin_america_codes,
  indicator = "NY.GDP.PCAP.PP.CD",
  start = 1995,
  end = 2022
)

# Clean and prepare regional data
gdp_regional_data <- gdp_regional_data %>%
  rename(
    year = year,
    country = country,
    gdp_ppp = NY.GDP.PCAP.PP.CD
  ) %>%
  filter(!is.na(gdp_ppp))

# Calculate regional average by year
regional_average <- gdp_regional_data %>%
  group_by(year) %>%
  summarise(
    gdp_ppp = mean(gdp_ppp, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA"  # Adding iso2c for consistency
  )

# Get data specifically for Argentina and Chile
gdp_focus_countries <- gdp_regional_data %>%
  filter(iso2c %in% c("AR", "CL"))

# Combine focus countries with regional average
gdp_combined <- bind_rows(gdp_focus_countries, regional_average)

# Create country labels for better display
gdp_combined <- gdp_combined %>%
  mutate(
    country_label = case_when(
      iso2c == "AR" ~ "Argentina",
      iso2c == "CL" ~ "Chile",
      country == "Regional Average" ~ "Media Regional (América Latina)",
      TRUE ~ country
    )
  )

# Plot the time series
ggplot(gdp_combined, aes(x = year, y = gdp_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
    title = "GDP per Capita (PPP, Constant International USD)",
    subtitle = "Argentina - Chile (1995–2022)",
    x = "Año",
    y = "PBI per Capita (USD, PPP)",
    color = "Country",
    caption = "Fuente: World Bank (WDI), indicator NY.GDP.PCAP.PP.CD"
  ) +
  scale_color_manual(values = colores_completos)+
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))

# Plot the time series (version for saving)
pp <- ggplot(gdp_combined, aes(x = year, y = gdp_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
    #  title = "GDP per Capita (PPP, Constant International USD)",
    #  subtitle = "Argentina & Chile (1995–2022)",
    x = "Año",
    y = "PBI per Capita (USD, PPP)",
    color = "País"
    #  caption = "Source: World Bank (WDI), indicator NY.GDP.PCAP.PP.CD"
  ) +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = colores_completos)+
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))

# Save plot
ggsave("output_context/gdp_per_capita_arg_chl_media.png", plot = pp, width = 8, height = 6)

```




**Calculo GDP otros paises para comparar**

```{r echo=FALSE, warning=FALSE}


    

# Get GDP data for all Latin American countries
gdp_regional_data <- WDI(
  country = latin_america_codes,
  indicator = "NY.GDP.PCAP.PP.CD",
  start = 1995,
  end = 2022
)

# Clean and prepare regional data
gdp_regional_data <- gdp_regional_data %>%
  rename(
    year = year,
    country = country,
    gdp_ppp = NY.GDP.PCAP.PP.CD
  ) %>%
  filter(!is.na(gdp_ppp))

# Calculate regional average by year
regional_average <- gdp_regional_data %>%
  group_by(year) %>%
  summarise(
    gdp_ppp = mean(gdp_ppp, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA"  # Adding iso2c for consistency
  )

# Get data specifically for Argentina and Chile
gdp_focus_countries <- gdp_regional_data %>%
  filter(iso2c %in% c("AR", "CL","BR","UY"))

# Combine focus countries with regional average
gdp_combined <- bind_rows(gdp_focus_countries, regional_average)

# Create country labels for better display
gdp_combined <- gdp_combined %>%
  mutate(
     country_label = case_when(
      iso2c == "AR" ~ "Argentina",
      iso2c == "CL" ~ "Chile",
      iso2c == "BR" ~ "Brasil",
      iso2c == "UY" ~ "Uruguay",
      country == "Regional Average" ~ "Media Regional (América Latina)",
      TRUE ~ country
      
    )
  )

# Plot the time series
ggplot(gdp_combined, aes(x = year, y = gdp_ppp, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = dollar_format(prefix = "USD ")) +
  labs(
    title = "GDP per Capita (PPP, Constant International USD)",
    subtitle = "Argentina & Chile (1995–2022)",
    x = "Year",
    y = "GDP per Capita (USD, PPP)",
    color = "Country",
    caption = "Source: World Bank (WDI), indicator NY.GDP.PCAP.PP.CD"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))


```


## GINI




```{r echo=FALSE, warning=FALSE,fig.width=8, fig.height=6}

# Get GDP data for all Latin American countries
gini_data <- WDI(
  country = latin_america_codes,
  indicator = "SI.POV.GINI",
  start = 1995,
  end = 2022
)

# Clean and prepare regional data
gini_data <- gini_data %>%
  rename(
    year = year,
    country = country,
    gini = SI.POV.GINI
  ) %>%
  filter(!is.na(gini))

# Calculate regional average by year
regional_average <- gini_data %>%
  group_by(year) %>%
  summarise(
    gini = mean(gini, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA"  # Adding iso2c for consistency
  )

# Get data specifically for Argentina and Chile
gini_focus_countries <- gini_data %>%
  filter(iso2c %in% c("AR", "CL"))

# Combine focus countries with regional average
gini_combined <- bind_rows(gini_focus_countries, regional_average)

# Create country labels for better display
gini_combined <- gini_combined %>%
  mutate(
    country_label = case_when(
      iso2c == "AR" ~ "Argentina",
      iso2c == "CL" ~ "Chile",
      country == "Regional Average" ~ "Media Regional (América Latina)",
      TRUE ~ country
    )
  )


# Plot the time series
ggplot(gini_combined, aes(x = year, y = gini, color = country_label, group = country_label)) +
   geom_line(size = 1.2) +
   geom_point(size = 2) +
   labs(
    title = "Coeficiente de GINI ",
    subtitle = "Argentina & Chile (1995–2022)",
    x = "Year",
    y = "Índice de Gini (0 = equidad perfecta, 100 = desigualdad total)",
    color = "Country",
    caption = "Source: World Bank (WDI), indicator SI.POV.GINI"
  ) +
  theme_minimal(base_size = 12) +
    scale_color_manual(values = colores_completos)+
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))



# Plot the time series (version for saving)
pp <- ggplot(gini_combined, aes(x = year, y = gini, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    #  title = "GDP per Capita (PPP, Constant International USD)",
    #  subtitle = "Argentina & Chile (1995–2022)",
    x = "Año",
    y = "Índice de Gini (0 = equidad perfecta, 100 = desigualdad total)",
    color = "País"
    #  caption = "Source: World Bank (WDI), indicator NY.GDP.PCAP.PP.CD"
  ) +
  theme_minimal(base_size = 12) +
    scale_color_manual(values = colores_completos)+
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 1))

# Save plot
ggsave("output_context/gini_arg_chl_media.png", plot = pp, width = 8, height = 6)


```




##Indicadores educativos



```{r}


# Define education indicators
education_indicators <- c(
  "SE.PRM.NENR",        # Primary education, pupils out-of-school (% of primary school age)
  "SE.SEC.NENR",        # Secondary education, pupils out-of-school (% of secondary school age)
  "SE.ADT.LITR.ZS",     # Literacy rate, adult total (% of people ages 15 and above)
  "SE.TER.ENRR",        # School enrollment, tertiary (% gross)
  "SE.PRM.CMPT.ZS",     # Primary completion rate, total (% of relevant age group)
  "SE.SEC.CMPT.LO.ZS",  # Lower secondary completion rate, total (% of relevant age group)
  "BAR.SCHL.15UP",      # Barro-Lee: Average years of schooling, age 15+, total
  "HD.HCI.EYRS",        # Expected years of school
  "SE.SCH.LIFE",        # School life expectancy (years)
  "SE.SEC.CUAT.UP.ZS",  # Educational attainment, at least completed upper secondary, population 25+, total (%) (cumulative)
  "SE.PRM.CUAT.ZS",     # Educational attainment, at least completed primary, population 25+ years, total (%) (cumulative)
  "HD.HCI.OVRL"         # Human Capital Index (HCI) (scale 0-1)
)

# Define year of interest
year_of_interest <- 2019

# Get education data
education_data <- WDI(
  country = latin_america_codes,
  indicator = education_indicators,
  start = 2019,
  end = 2019
)

# Clean and prepare data
education_data_clean <- education_data %>%
  filter(year == year_of_interest) %>%
  select(iso2c, country, year, all_of(education_indicators)) %>%
  # Rename indicators for better readability
  rename(
    primary_out_of_school = SE.PRM.NENR,
    secondary_out_of_school = SE.SEC.NENR,
    adult_literacy = SE.ADT.LITR.ZS,
    tertiary_enrollment = SE.TER.ENRR,
    primary_completion = SE.PRM.CMPT.ZS,
    secondary_completion = SE.SEC.CMPT.LO.ZS,
    avg_years_schooling = BAR.SCHL.15UP,
    expected_years_school = HD.HCI.EYRS,
    school_life_expectancy = SE.SCH.LIFE,
    upper_secondary_attainment = SE.SEC.CUAT.UP.ZS,
    primary_attainment = SE.PRM.CUAT.ZS,
    human_capital_index = HD.HCI.OVRL
  ) %>%
  # Calcular tasa de analfabetismo
  mutate(
    illiteracy_rate = 100 - adult_literacy
  )

# Calculate Latin America averages
latin_america_averages <- education_data_clean %>%
  summarise(
    across(c(primary_out_of_school, secondary_out_of_school, adult_literacy, 
             tertiary_enrollment, primary_completion, secondary_completion,
             avg_years_schooling, expected_years_school, school_life_expectancy,
             upper_secondary_attainment, primary_attainment, human_capital_index,
             illiteracy_rate), 
           ~ mean(.x, na.rm = TRUE)),
    .groups = 'drop'
  ) %>%
  mutate(
    iso2c = "LA",
    country = "América Latina (Promedio)",
    year = year_of_interest
  )

# Get data for Argentina and Chile
focus_countries_data <- education_data_clean %>%
  filter(iso2c %in% c("AR", "CL"))

# Combine all data
combined_data <- bind_rows(focus_countries_data, latin_america_averages) %>%
  arrange(country)

# Create summary table
education_table <- combined_data %>%
  select(-year, -iso2c) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  # Transpose to have indicators as rows
  pivot_longer(
    cols = -country,
    names_to = "Indicador",
    values_to = "Valor"
  ) %>%
  pivot_wider(
    names_from = country,
    values_from = Valor
  ) %>%
  mutate(
    Indicador = case_when(
      Indicador == "primary_out_of_school" ~ "Niños fuera de escuela primaria (%)",
      Indicador == "secondary_out_of_school" ~ "Niños fuera de escuela secundaria (%)",
      Indicador == "adult_literacy" ~ "Tasa de alfabetización adultos (%)",
      Indicador == "illiteracy_rate" ~ "Tasa de analfabetismo (%)",
      Indicador == "tertiary_enrollment" ~ "Matrícula educación terciaria (%)",
      Indicador == "primary_completion" ~ "Tasa finalización primaria (%)",
      Indicador == "secondary_completion" ~ "Tasa finalización sec. inferior (%)",
      Indicador == "avg_years_schooling" ~ "Años promedio de escolaridad (15+)",
      Indicador == "expected_years_school" ~ "Años esperados de escolaridad",
      Indicador == "school_life_expectancy" ~ "Esperanza de vida escolar (años)",
      Indicador == "upper_secondary_attainment" ~ "Población 25+ con secundaria superior completa (%)",
      Indicador == "primary_attainment" ~ "Población 25+ con primaria completa (%)",
      Indicador == "human_capital_index" ~ "Índice de Capital Humano (0-1)",
      TRUE ~ Indicador
    )
  )

# Print table
cat("=== INDICADORES EDUCATIVOS 2020 ===\n\n")
cat("Argentina, Chile y América Latina\n\n")

table_output <- kable(education_table, 
              format = "markdown", 
              align = c("l", rep("c", ncol(education_table)-1)),
              caption = "Indicadores Educativos - 2019")

print(table_output)



```


```{r}

# Datos manuales de alfabetización para Argentina y Chile (desde 1995)
manual_data <- data.frame(
  year = c(1996, 1998, 2000, 2003, 2006, 2009, 2011, 2013, 2015, 2017, 2022),
  CH_literacy = c(95.9, 96.2, 96.6, 96.6, 97.0, 97.4, 97.5, 97.1, 97.8, 97.2, 98.2),
  AR_literacy = c(98.8, 98.7, 98.6, 98.8, 99.0, 98.9, 99.3, 99.3, 99.3, 99.3, 99.3)
)

# Convertir a formato largo y calcular analfabetismo
manual_data_long <- manual_data %>%
  pivot_longer(
    cols = c(CH_literacy, AR_literacy),
    names_to = "country",
    values_to = "literacy_rate"
  ) %>%
  mutate(
    illiteracy_rate = 100 - literacy_rate,
    iso2c = case_when(
      country == "CH_literacy" ~ "CL",
      country == "AR_literacy" ~ "AR"
    ),
    country_label = case_when(
      country == "CH_literacy" ~ "Chile",
      country == "AR_literacy" ~ "Argentina"
    )
  )

# Get literacy data for all Latin American countries to calculate regional average
literacy_regional_data <- WDI(
  country = latin_america_codes,
  indicator = "SE.ADT.LITR.ZS",
  start = 1995,
  end = 2022
)

# Calculate regional average by year (excluding 1994, starting from 1995)
regional_average <- literacy_regional_data %>%
  rename(
    literacy_rate = SE.ADT.LITR.ZS
  ) %>%
  filter(!is.na(literacy_rate)) %>%
  filter(year >= 1995) %>%  # EMPEZAR DESDE 1995
  mutate(
    illiteracy_rate = 100 - literacy_rate
  ) %>%
  group_by(year) %>%
  summarise(
    illiteracy_rate = mean(illiteracy_rate, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    country = "Regional Average",
    iso2c = "LA",
    country_label = "Media Regional (América Latina)"
  )

# Combine manual data with regional average
illiteracy_combined <- bind_rows(
  manual_data_long %>% select(year, illiteracy_rate, iso2c, country_label),
  regional_average %>% select(year, illiteracy_rate, iso2c, country_label)
)

# Plot the time series
pp <- ggplot(illiteracy_combined, aes(x = year, y = illiteracy_rate, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1995, 2022, by = 3)) +
    scale_color_manual(values = colores_completos)+
  labs(
    title = "Tasa de Analfabetismo (% población 15+ años)",
    subtitle = "Argentina, Chile y Media Regional de América Latina (1995-2022)",
    x = "Año",
    y = "Tasa de Analfabetismo (%)",
    color = "País/Región",
    caption = "Fuente: UNESCO Institute for Statistics (Argentina y Chile)- Banco Mundial - WDI (Media Regional) \ SEDLAC (CEDLAS y Banco Mundial)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  guides(color = guide_legend(nrow = 1))

# Show plot
print(pp)

# Plot the time series
pp2 <- ggplot(illiteracy_combined, aes(x = year, y = illiteracy_rate, color = country_label, group = country_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(1995, 2022, by = 3)) +
    scale_color_manual(values = colores_completos)+
  labs(
#    title = "Tasa de Analfabetismo (% población 15+ años)",
#    subtitle = "Argentina, Chile y Media Regional de América Latina (1995-2022)",
    x = "Año",
    y = "Tasa de Analfabetismo (%)",
    color = "País/Región"
#    caption = "Fuente: UNESCO Institute for Statistics (Argentina y Chile)\Banco Mundial - WDI (Media Regional) \ SEDLAC (CEDLAS y Banco #Mundial)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  guides(color = guide_legend(nrow = 1))


# Save plot
ggsave("output_context/tasa_analfabetismo_arg_chl_media.png", plot = pp2, width = 10, height = 6, dpi = 300)


```



```{r echo=FALSE, warning=FALSE}

library(migest)
library(countrycode)
library(sf)
library(scales)

rm(list = ls())

## Base UNU
unu<- read_csv("Data/migration_imputed_RIKS_dec2021.csv")

#suramerica
countries_south_america <- c(
  'ARG', 'BOL', 'BRA', 'CHL', 'COL', 'ECU',
  'GUY', 'PRY', 'PER', 'SUR', 'URY', 'VEN'
)
countries_latam <- c(
  # Centroamérica continental
  'BLZ', 'CRI', 'SLV', 'GTM', 'HND', 'NIC', 'PAN', 'MEX',
  # Caribe insular
  'CUB', 'DOM', 'HTI', 'JAM', 'TTO', 'BRB', 'BHS', 'GRD', 'LCA', 'VCT', 'ATG', 'KNA', 'DMA', 'SLB',
  'ARG', 'BOL', 'BRA', 'CHL', 'COL', 'ECU',
  'GUY', 'PRY', 'PER', 'SUR', 'URY', 'VEN'
)

countries_north_america <- c('USA', 'CAN')

countries_arg_chl <- c('ARG', 'CHL')


```

**Construiyo migrantes agrupados por pais de destino y corrigo daros arg y chl**


```{r}
# Base agrupada por país de destino y año
unu_country_dest <- unu %>%
  group_by(iso_des, year) %>%
  summarise(stock = sum(stock, na.rm = TRUE),
            flow = sum(flow, na.rm = TRUE),
            inflow = sum(inflow, na.rm = TRUE),
            outflow = sum(outflow, na.rm = TRUE)) %>%
  ungroup()

# Cargar datos ajustados
df_unu_adjust <- read_csv("data/UNU_CRIS_adjusted.csv")
df_unu_adjust <- df_unu_adjust %>%
  filter(iso_des == "ARG" | iso_des == "CHL") 

# SOLUCIÓN: Eliminar ARG y CHL de la base original y luego agregar los ajustados
unu_country_dest <- unu_country_dest %>%
  filter(!(iso_des %in% c("ARG", "CHL"))) %>%  # Eliminar ARG y CHL originales
  bind_rows(df_unu_adjust) %>%  # Agregar datos ajustados completos
  arrange(iso_des, year)  # Ordenar para mantener estructura

# Filtrar para los países de América Latina
unu_latam <- unu_country_dest %>%
  filter(iso_des %in% countries_latam)

unu_arg_chl <- unu_latam %>%
  filter(iso_des %in% c("ARG", "CHL"))


# Paleta homogénea (del análisis ISDM)
colores_pais <- c(
  "CHL" = "#1BA3A3",       
  "ARG" = "#D73027"    
)

colores_completos <- c(
  "ARG" = "#D73027",
  "CHL" = "#1BA3A3",
  "Media Regional (América Latina)" = "#FFA726"
)

```

## Stock migratorio de Argentina y Chile


```{r}

# Separar datos en dos períodos
unu_antes_2020 <- unu_arg_chl %>% filter(year < 2020)
unu_desde_2020 <- unu_arg_chl %>% filter(year >= 2020)

# Preparar datos para etiquetas 2024
datos_2024 <- unu_arg_chl %>%
  filter(year == 2024)

# Gráfico de líneas con distinción de datos ajustados
ggplot(unu_arg_chl, aes(x = year, y = stock/1000, color = iso_des, group = iso_des)) +
  # Líneas sólidas para antes de 2020
  geom_line(
    data = unu_antes_2020,
    linewidth = 0.8,
    alpha = 0.9
  ) +
  # Líneas punteadas para 2020-2024
  geom_line(
    data = unu_desde_2020,
    linetype = "dotted",
    linewidth = 0.8,
    alpha = 0.9
  ) +
  # Más breaks en el eje Y
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 8),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  # Más breaks en el eje X incluyendo 2024
  scale_x_continuous(
    breaks = seq(min(unu_arg_chl$year), 2024, by = 4)
  ) +
  labs(
    title = "Stock migratorio de Argentina y Chile",
    subtitle = "Líneas punteadas: datos ajustados (2020-2024)",
    x = "Año",
    y = "Stock (x 1.000)",
    color = "País"
  ) +
  theme_minimal(base_size = 13) +
    scale_color_manual(values = colores_completos)+
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40"),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "gray90", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(nrow = 1, override.aes = list(linewidth = 3)))

#guardo sin titulo para el documento
# Gráfico de líneas con distinción de datos ajustados
pp<-ggplot(unu_arg_chl, aes(x = year, y = stock/1000, color = iso_des, group = iso_des)) +
  # Líneas sólidas para antes de 2020
  geom_line(
    data = unu_antes_2020,
    linewidth = 0.8,
    alpha = 0.9
  ) +
  # Líneas punteadas para 2020-2024
  geom_line(
    data = unu_desde_2020,
    linetype = "dotted",
    linewidth = 0.8,
    alpha = 0.9
  ) +
  # Más breaks en el eje Y
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 8),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  # Más breaks en el eje X incluyendo 2024
  scale_x_continuous(
    breaks = seq(min(unu_arg_chl$year), 2024, by = 4)
  ) +
  labs(
    x = "Año",
    y = "Stock (x 1.000)",
    color = "País"
  ) +
  theme_minimal(base_size = 13) +
   scale_color_manual(values = colores_completos)+
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40"),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "gray90", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.text = element_text(color = "black"),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(nrow = 1, override.aes = list(linewidth = 3)))

#save plot pp
ggsave("output_context/stock_migratorio_arg_chl.png", plot = pp, width = 10, height = 6, dpi = 300)

```

