---
title: "ISDM — Integrated Analysis: Migrants and Recent Migrants"
subtitle: "Gaps by Dimension and Synthetic Migration Inequality Index — Argentina and Chile 2015-2024"
author: "Natalia Debandi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 7
)
```

```{r libs}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gt)
```

# Datos

```{r carga}
# brecha_nat_mig: CSV con ENE (más denso para indicadores laborales)
# brecha_nat_rec: CSV solo CASEN (fuente única para migrantes recientes)
indicadores_full <- read_csv(
  "Data/indicadores_con_brechas_arg_chile_2015_2024.csv",
  show_col_types = FALSE
)
indicadores_sc <- read_csv(
  "Data/indicadores_con_brechas_solocasen_arg_chile_2015_2024.csv",
  show_col_types = FALSE
)

indicadores_con_brechas <- indicadores_full %>%
  select(pais, fuente, anio, indicador, indicador_clean, brecha_nat_mig) %>%
  left_join(
    indicadores_sc %>% select(pais, anio, indicador, brecha_nat_rec),
    by = c("pais", "anio", "indicador")
  )

cat("Rows:", nrow(indicadores_con_brechas), "\n")
cat("Period:", min(indicadores_con_brechas$anio), "-", max(indicadores_con_brechas$anio), "\n")
cat("Indicators:", paste(sort(unique(indicadores_con_brechas$indicador)), collapse = ", "), "\n")
```

```{r clasificacion}
clasificacion_indicadores <- tribble(
  ~indicador,                     ~dimension,  ~tipo,
  "I1_tasa_actividad",            "Labor",   "positivo",
  "I2_tasa_desempleo",            "Labor",   "negativo",
  "I3_informalidad_asalariada",   "Labor",   "negativo",
  "I4_cuentapropismo",            "Labor",   "negativo",
  "I5_pobreza",                   "Poverty",   "negativo",
  "I6_pobreza_extrema",           "Poverty",   "negativo",
  "I7_asistencia_escolar_6_17",   "Social",    "positivo",
  "I8_asistencia_superior_18_29", "Social",    "positivo",
  "I9_sistema_salud_publico",     "Social",    "positivo"
)

colores_tipo <- c(
  "Migrants"           = "#D95F02",
  "Recent migrants" = "#7570B3"
)

colores_pais <- c(
  "Argentina" = "#2E86AB",
  "Chile"     = "#A23B72"
)

etiquetas_ind <- c(
  "I1_tasa_actividad"            = "Activity rate",
  "I2_tasa_desempleo"            = "Unemployment rate",
  "I3_informalidad_asalariada"   = "Salaried informality",
  "I4_cuentapropismo"            = "Self-employment",
  "I5_pobreza"                   = "Monetary poverty",
  "I6_pobreza_extrema"           = "Extreme poverty",
  "I7_asistencia_escolar_6_17"   = "School attendance (6-17)",
  "I8_asistencia_superior_18_29" = "Higher education attendance (18-29)",
  "I9_sistema_salud_publico"     = "Public health access"
)

if (!dir.exists("output_integrados")) dir.create("output_integrados")
```

```{r preparacion}
# Formato largo: una fila por indicador × año × país × tipo_migrante
datos_largo <- indicadores_con_brechas %>%
  filter(indicador %in% clasificacion_indicadores$indicador) %>%
  select(pais, fuente, anio, indicador, indicador_clean,
         brecha_nat_mig, brecha_nat_rec) %>%
  pivot_longer(
    cols      = c(brecha_nat_mig, brecha_nat_rec),
    names_to  = "tipo_migrante",
    values_to = "brecha"
  ) %>%
  mutate(
    tipo_migrante = if_else(tipo_migrante == "brecha_nat_mig",
                            "Migrants", "Recent migrants")
  ) %>%
  left_join(clasificacion_indicadores, by = "indicador") %>%
  # Normalización direccional: brecha > 0 siempre = nativos mejor
  mutate(
    brecha_normalizada = if_else(tipo == "negativo", -brecha, brecha)
  )

# Estandarización conjunta: media/SD calculados sobre ambos tipos de migrante
datos_largo <- datos_largo %>%
  group_by(indicador) %>%
  mutate(
    media_conj = mean(brecha_normalizada, na.rm = TRUE),
    sd_conj    = sd(brecha_normalizada,   na.rm = TRUE),
    brecha_z   = (brecha_normalizada - media_conj) / sd_conj
  ) %>%
  ungroup()

cat("✓ Data prepared:", nrow(datos_largo), "rows\n")
cat("  Migrant types:", paste(unique(datos_largo$tipo_migrante), collapse = ", "), "\n")
```

---

# Labor Dimension

> **Interpretation note:** normalized gap = Natives − Migrants (sign-adjusted). Positive values indicate natives are better off; negative values indicate migrants are better off. Each point represents a year.

## Without Standardization (percentage points)

```{r laboral-raw, fig.height=6}
p <- datos_largo %>%
  filter(dimension == "Labor", !is.na(brecha_normalizada)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_normalizada, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Normalized gap (pp)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/laboral_original.png", plot = p, width = 14, height = 6, dpi = 300)
```

## Joint Standardization (Z-scores)

Standardization is computed over the **joint distribution** of both groups (migrants and recent migrants), so Z-scores are directly comparable across groups.

$$Z_i = \frac{x_i - \bar{x}_{joint}}{\sigma_{joint}}$$

where $\bar{x}_{joint}$ and $\sigma_{joint}$ are computed over all observations of the indicator (both migrant types, both countries, all years).

```{r laboral-z, fig.height=6}
p <- datos_largo %>%
  filter(dimension == "Labor", !is.na(brecha_z)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_z, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Standardized gap (SD)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/laboral_z.png", plot = p, width = 14, height = 6, dpi = 300)
```

---

# Poverty Dimension

## Without Standardization (percentage points)

```{r pobreza-raw, fig.height=5}
p <- datos_largo %>%
  filter(dimension == "Poverty", !is.na(brecha_normalizada)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_normalizada, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Normalized gap (pp)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/pobreza_original.png", plot = p, width = 14, height = 5, dpi = 300)
```

## Joint Standardization (Z-scores)

```{r pobreza-z, fig.height=5}
p <- datos_largo %>%
  filter(dimension == "Poverty", !is.na(brecha_z)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_z, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Standardized gap (SD)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/pobreza_z.png", plot = p, width = 14, height = 5, dpi = 300)
```

---

# Social Dimension

## Without Standardization (percentage points)

```{r social-raw, fig.height=6}
p <- datos_largo %>%
  filter(dimension == "Social", !is.na(brecha_normalizada)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_normalizada, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Normalized gap (pp)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/social_original.png", plot = p, width = 14, height = 6, dpi = 300)
```

## Joint Standardization (Z-scores)

```{r social-z, fig.height=6}
p <- datos_largo %>%
  filter(dimension == "Social", !is.na(brecha_z)) %>%
  mutate(
    ind_num         = as.integer(str_extract(indicador, "(?<=I)\\d+")),
    indicador_label = paste0(str_extract(indicador, "I\\d+"), " - ", etiquetas_ind[indicador]),
    indicador_label = fct_reorder(indicador_label, ind_num)
  ) %>%
  ggplot(aes(x = indicador_label, y = brecha_z, color = pais)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  geom_boxplot(alpha = 0.5, position = position_dodge(0.8), outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 2) +
  coord_flip() +
  scale_color_manual(values = colores_pais, name = "Country") +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Standardized gap (SD)\nLeft (negative) = Migrants better | Right (positive) = Natives better",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold", size = 13),
    axis.text        = element_text(size = 11, color = "black"),
    axis.title       = element_text(size = 12, color = "black"),
    panel.grid.minor = element_blank()
  )
print(p)
ggsave("output_integrados/social_z.png", plot = p, width = 14, height = 6, dpi = 300)
```

---

# Summary Tables

> **Note:** values expressed in percentage points (pp). Positive = natives better off; negative = migrants better off. Mean calculated over all available years.

```{r tablas-prep}
resumen <- datos_largo %>%
  filter(!is.na(brecha_normalizada), anio != 2020) %>%
  group_by(dimension, indicador, pais, tipo_migrante) %>%
  summarise(media = mean(brecha_normalizada, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from  = c(tipo_migrante, pais),
    values_from = media,
    names_sep   = "_"
  ) %>%
  mutate(
    dif_mig     = `Migrants_Chile` - `Migrants_Argentina`,
    dif_rec     = `Recent migrants_Chile` - `Recent migrants_Argentina`,
    ind_label   = etiquetas_ind[indicador]
  )
```

## Labor Dimension

```{r tabla-laboral}
resumen %>%
  filter(dimension == "Labor") %>%
  select(
    Indicador = ind_label,
    mig_ARG   = `Migrants_Argentina`,
    mig_CHL   = `Migrants_Chile`,
    mig_dif   = dif_mig,
    rec_ARG   = `Recent migrants_Argentina`,
    rec_CHL   = `Recent migrants_Chile`,
    rec_dif   = dif_rec
  ) %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 1) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 General Migrants (pp)",
    columns = c(mig_ARG, mig_CHL, mig_dif)
  ) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 Recent Migrants (pp)",
    columns = c(rec_ARG, rec_CHL, rec_dif)
  ) %>%
  cols_label(
    mig_ARG = "ARG", mig_CHL = "CHL", mig_dif = "CHL\u2212ARG",
    rec_ARG = "ARG", rec_CHL = "CHL", rec_dif = "CHL\u2212ARG"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  )
```

## Poverty Dimension

```{r tabla-pobreza}
resumen %>%
  filter(dimension == "Poverty") %>%
  select(
    Indicador = ind_label,
    mig_ARG   = `Migrants_Argentina`,
    mig_CHL   = `Migrants_Chile`,
    mig_dif   = dif_mig,
    rec_ARG   = `Recent migrants_Argentina`,
    rec_CHL   = `Recent migrants_Chile`,
    rec_dif   = dif_rec
  ) %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 1) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 General Migrants (pp)",
    columns = c(mig_ARG, mig_CHL, mig_dif)
  ) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 Recent Migrants (pp)",
    columns = c(rec_ARG, rec_CHL, rec_dif)
  ) %>%
  cols_label(
    mig_ARG = "ARG", mig_CHL = "CHL", mig_dif = "CHL\u2212ARG",
    rec_ARG = "ARG", rec_CHL = "CHL", rec_dif = "CHL\u2212ARG"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  )
```

## Social Dimension

```{r tabla-social}
resumen %>%
  filter(dimension == "Social") %>%
  select(
    Indicador = ind_label,
    mig_ARG   = `Migrants_Argentina`,
    mig_CHL   = `Migrants_Chile`,
    mig_dif   = dif_mig,
    rec_ARG   = `Recent migrants_Argentina`,
    rec_CHL   = `Recent migrants_Chile`,
    rec_dif   = dif_rec
  ) %>%
  gt() %>%
  fmt_number(columns = where(is.numeric), decimals = 1) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 General Migrants (pp)",
    columns = c(mig_ARG, mig_CHL, mig_dif)
  ) %>%
  tab_spanner(
    label   = "Gap: Native \u2212 Recent Migrants (pp)",
    columns = c(rec_ARG, rec_CHL, rec_dif)
  ) %>%
  cols_label(
    mig_ARG = "ARG", mig_CHL = "CHL", mig_dif = "CHL\u2212ARG",
    rec_ARG = "ARG", rec_CHL = "CHL", rec_dif = "CHL\u2212ARG"
  ) %>%
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  )
```

---

# Integrated ISDM

> Z-scores calculated **separately** for general migrants and recent migrants (consistent with individual reports). Year 2020 is excluded in both groups. Sources: `brecha_nat_mig` from the ENE CSV; `brecha_nat_rec` from the CASEN-only CSV.

## Calculation

```{r isdm-prep}
# Estandarización por indicador × tipo_migrante (separada, igual que reportes individuales)
datos_isdm <- datos_largo %>%
  filter(anio != 2020, !is.na(brecha_normalizada)) %>%
  group_by(indicador, tipo_migrante) %>%
  mutate(
    media_ind    = mean(brecha_normalizada, na.rm = TRUE),
    sd_ind       = sd(brecha_normalizada,   na.rm = TRUE),
    brecha_z_ind = (brecha_normalizada - media_ind) / sd_ind
  ) %>%
  ungroup()

# Sub-índices por dimensión
subindices_integrado <- datos_isdm %>%
  group_by(pais, anio, dimension, tipo_migrante) %>%
  summarise(
    sub_abs_norm = mean(abs(brecha_normalizada), na.rm = TRUE),
    sub_abs_std  = mean(abs(brecha_z_ind),       na.rm = TRUE),
    sub_dir_norm = mean(brecha_normalizada,       na.rm = TRUE),
    sub_dir_std  = mean(brecha_z_ind,             na.rm = TRUE),
    .groups = "drop"
  )

# ISDM final: promedio jerárquico sobre dimensiones
isdm_integrado <- subindices_integrado %>%
  group_by(pais, anio, tipo_migrante) %>%
  summarise(
    isdm_abs_norm = mean(sub_abs_norm, na.rm = TRUE),
    isdm_abs_std  = mean(sub_abs_std,  na.rm = TRUE),
    isdm_dir_norm = mean(sub_dir_norm, na.rm = TRUE),
    isdm_dir_std  = mean(sub_dir_std,  na.rm = TRUE),
    .groups = "drop"
  )

cat("✓ Integrated ISDM calculated\n")
cat("  Years:", paste(sort(unique(isdm_integrado$anio)), collapse = ", "), "\n")
cat("  Groups:", paste(unique(isdm_integrado$tipo_migrante), collapse = ", "), "\n")
```

## Comparative Evolution — Absolute ISDM (Standardized)

```{r isdm-evolucion-std, fig.height=6}
anios_breaks_isdm <- sort(unique(isdm_integrado$anio))

p_isdm_std <- ggplot(isdm_integrado,
       aes(x = anio, y = isdm_abs_std,
           color = pais, linetype = tipo_migrante)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3, alpha = 0.8) +
  scale_x_continuous(breaks = anios_breaks_isdm) +
  scale_color_manual(values = colores_pais, name = "Country") +
  scale_linetype_manual(
    values = c("Migrants" = "solid", "Recent migrants" = "dashed"),
    name = "Grupo"
  ) +
  labs(
    x = "Year",
    y = "Inequality magnitude (SD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_isdm_std)
ggsave("output_integrados/isdm_evolucion_std.png",
       plot = p_isdm_std, width = 12, height = 6, dpi = 300)
```

## Comparative Evolution — Absolute ISDM (Normalized, pp)

```{r isdm-evolucion-pp, fig.height=6}
p_isdm_pp <- ggplot(isdm_integrado,
       aes(x = anio, y = isdm_abs_norm,
           color = pais, linetype = tipo_migrante)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3, alpha = 0.8) +
  scale_x_continuous(breaks = anios_breaks_isdm) +
  scale_color_manual(values = colores_pais, name = "Country") +
  scale_linetype_manual(
    values = c("Migrants" = "solid", "Recent migrants" = "dashed"),
    name = "Grupo"
  ) +
  labs(
    x = "Year",
    y = "Inequality magnitude (pp)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_isdm_pp)
ggsave("output_integrados/isdm_evolucion_pp.png",
       plot = p_isdm_pp, width = 12, height = 6, dpi = 300)
```

## Evolution by Country — Absolute ISDM (Standardized)

```{r isdm-evolucion-facet, fig.height=5}
p_isdm_facet <- ggplot(isdm_integrado,
       aes(x = anio, y = isdm_abs_std,
           color = tipo_migrante, linetype = tipo_migrante)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3, alpha = 0.8) +
  scale_x_continuous(breaks = anios_breaks_isdm) +
  scale_color_manual(
    values = c("Migrants" = "#D95F02", "Recent migrants" = "#7570B3"),
    name = "Grupo"
  ) +
  scale_linetype_manual(
    values = c("Migrants" = "solid", "Recent migrants" = "dashed"),
    name = "Grupo"
  ) +
  facet_wrap(~pais, ncol = 2) +
  labs(
    x = "Year",
    y = "Inequality magnitude (SD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

print(p_isdm_facet)
ggsave("output_integrados/isdm_evolucion_facet.png",
       plot = p_isdm_facet, width = 12, height = 5, dpi = 300)
```

## Direction of Inequality by Dimension (pp)

```{r isdm-direccional, fig.height=5}
resumen_dim_integrado <- subindices_integrado %>%
  group_by(pais, dimension, tipo_migrante) %>%
  summarise(
    subindice_dir_promedio = mean(sub_dir_norm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(dimension = factor(dimension, levels = c("Labor", "Poverty", "Social")))

p_dir_integrado <- ggplot(resumen_dim_integrado,
       aes(x = dimension, y = subindice_dir_promedio, fill = pais)) +
  geom_hline(yintercept = 0, linewidth = 1, color = "black") +
  geom_col(position = "dodge", width = 0.7, alpha = 0.9) +
  geom_text(
    aes(
      label = sprintf("%+.2f", subindice_dir_promedio),
      hjust = ifelse(subindice_dir_promedio > 0, -0.2, 1.2)
    ),
    position  = position_dodge(width = 0.7),
    size      = 3.5,
    fontface  = "bold"
  ) +
  scale_fill_manual(values = colores_pais, name = "Country") +
  coord_flip() +
  facet_wrap(~tipo_migrante, ncol = 2) +
  labs(
    x = "",
    y = "Directional sub-index (pp)\nNegative = Migrants better | Positive = Natives better"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

print(p_dir_integrado)
ggsave("output_integrados/isdm_direccional_dimensiones.png",
       plot = p_dir_integrado, width = 12, height = 5, dpi = 300)
```
